On exponential sums arising from the classical doubling method

### 2025-04-10 main.tex
n => N

The main result of this paper is 
\begin{theorem}
\label{thm:Phi_stable}
	The assignments $H \to \cComplex$ given by $$h \mapsto \genHermitianJacobiKernel{\hermitianSpace}{\chi}\left(h\right)$$ are stable functions. Equivalently, the assignments $$\pi \mapsto \dblJacobiSumScalar{\pi}{\chi}$$ are constant on geometric Lusztig series.
\end{theorem}
%In the $\GL_n$ case, \Cref{thm:gln-doubling-gauss-sum-in-terms-of-kondo}

\begin{lemma}
\label{lem:indep_geo_conj}
    If $(\algebraicGroup{T},\theta)$ and $(\algebraicGroup{T}', \theta)$ are geometrically conjugate, then $$\frac{\dblVirtualJacobiSumScalar{\RTTheta{T}{\theta}}{\chi}}{\RTTheta{T}{\theta}(1)} = \frac{\dblVirtualJacobiSumScalar{\RTTheta{T'}{\theta'}}{\chi}}{\RTTheta{T'}{\theta'}\left(1\right)}.$$
\end{lemma}

We recall that \elad{Maybe we can help the reader and tell them that $m=1,2$} $$
	\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n} = \prod_{a \in \mathbb{Z}/m\mathbb{Z}} \left(\algebraicGroup{\GL}_n\right)_a,
$$
with Frobenius root action $$\Frobenius(g_0,g_1, \hdots, g_{m-1}) = (\Frobenius(g_{m-1}), \Frobenius(g_0), \Frobenius(g_1), \hdots, \Frobenius(g_{m-2})).$$ We will freely identify $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n}(\finiteField) = \GL_n(\finiteFieldExtension{m})$ via the map $\left(g_0,\dots,g_{m-1}\right) \mapsto g_0$.

\begin{remark}
Torus character pairs for $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n}$ are the same as torus character pairs for $\algebraicGroup{\GL}_n$ as a group over $\finiteFieldExtension{m}$.
\end{remark}

@@

Given a torus character pair $(\algebraicGroup{T},\theta)$ with associated $\Frobenius$-twisted conjugacy class $(w) \in W(\algebraicGroup{T}_0)$ along with an embedding $\rho:\algebraicGroup{G}^* \to \restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n}$ and a Frobenius stable maximal torus $\algebraicGroup{T}^* \subset \restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n}$ such that $\rho(\algebraicGroup{T}_0^*) \subset \algebraicGroup{T}^*$, let $\tilde{\algebraicGroup{T}}$ be the torus of $\algebraicGroup{\GL}_n$ \elad{of $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_n}$?} corresponding to the $\Frobenius$-twisted conjugacy class of $\rho(w) \in W(\algebraicGroup{T})$ \elad{of $W\left(\algebraicGroup{T}^{\ast}\right)$? Or maybe $W\left(\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{A_n}\right)$?} and let $\tilde{\theta} \colon \tilde{\algebraicGroup{T}}^{\Frobenius} \to \multiplicativegroup{\cComplex}$ be a character such that $(\rho(s_{\algebraicGroup{T},\theta})) = (s_{\tilde{\algebraicGroup{T}},\tilde{\theta}})$. We define
\[
    g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi) := \tau_{\Tilde{\algebraicGroup{T}}}(\tilde{\theta} \times \chi \circ \det,\psi) = \epsilon_{\Tilde{\algebraicGroup{T}}} q^{-nm/2}\sum_{t \in \tilde{\algebraicGroup{T}}^{\Frobenius}} \tilde{\theta}^{-1}(t)\chi^{-1}({\det}_{\finiteFieldExtension{m}}t) \psi(\trace_{\finiteFieldExtension{m}}t).
\]
Here $\chi \colon \multiplicativegroup{\finiteFieldExtension{m}} \to \multiplicativegroup{\cComplex}$ and $\fieldCharacter \colon \finiteFieldExtension{m} \to \multiplicativegroup{\cComplex}$ are characters. The following lemma shows that this is well defined.

\begin{lemma}
\label{lemma:gl_invariance}
\begin{enumerate}[(i)]
	\item \label{item:gl-n-hasse-davenport-gauss-sum-implies-well-defined} In the case that $\rho$ is the identity map, $g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi)$ depends only on the semisimple conjugacy class of $(s_{\algebraicGroup{T},\theta})$ and the character $\chi$.
	\item The value $g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi)$ is well defined, i.e., it is independent of the choices $\algebraicGroup{T}_0$, $\algebraicGroup{T}^*$ and of the conjugacy class representative $w$.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}[(i)]
	\item In this case when $T \cong \multiplicativegroup{\finiteFieldExtension{\lambda}}$, we recover the Gauss sum
	\[
	g_{T,\rho}(\chi,\theta,\psi) = \GaussSumTorusCharacter{T}{\theta}{\chi}{\fieldCharacter} = \left(-1\right)^n \GaussSumScalar{\tau \times \chi}{\fieldCharacter},
	\]
	where $\tau$ is any irreducible subrepresentation of $\RTThetaVirtual{T}{\GL_n\left(\finiteField\right)}{\theta}$. By \Cref{thm:KondoClassical}, this only depends on the geometric conjugacy class of $(T,\theta)$ and hence only on the semisimple conjugacy class of $(s_{T,\theta})$.
	\item For a Frobenius stable semisimple conjugacy class $(s)$ in $\GL_n$, we define
	$$
	g_{(s)}(\chi,\fieldCharacter) := g_{T,\rho}(\chi,\theta,\fieldCharacter)$$
	where $(T,\theta)$ is a torus character pair such that $\left(s_{T,\theta}\right) = (s)$.
	By (\ref{item:gl-n-hasse-davenport-gauss-sum-implies-well-defined}) this is well defined. Then
	$$
	g_{T,\rho}(\chi,\theta,\fieldCharacter)  = g_{(\rho(s_{T,\theta}))}(\chi,\fieldCharacter),
	$$
	which is independent of the choices $\algebraicGroup{T}^* \subset \algebraicGroup{G}^*$ and $\algebraicGroup{T} \subset \algebraicGroup{\GL}_n$.
\end{enumerate}
\end{proof}

@@

The idea behind the proof of \Cref{lem:indep_geo_conj} is that the formula from \Cref{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} relates $\dblVirtualJacobiSumScalar{\RTTheta{T}{\theta}}{\chi}/\RTTheta{T}{\theta}\left(1\right)$ to $g_{T,\rho}(\chi,\theta,\psi)$ for the appropriate \calvin{doubling representation} $\rho$ in the table below.

\begin{center}
	\begin{tabular}{|c|c|} \hline
		$\algebraicGroup{G}$ & $\rho \colon \algebraicGroup{G}^{\ast} \to \left(\restrictionOfScalars{\quadraticExtension}{\finiteField}{}\right) \algebraicGroup{\GL}_N$  \tabularnewline \hline \hline
		$\algebraicGroup{GL}_n$ &  $\algebraicGroup{\GL}_n \to \algebraicGroup{GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\UnitaryGroup}_{n}$ &  $\algebraicGroup{\UnitaryGroup}_{n} \to \restrictionOfScalars{\quadraticExtension}{\finiteField}{\algebraicGroup{GL}_n}$\tabularnewline \hline		 				 
		$\algebraicGroup{\SO}_{2n+1}$ &$\algebraicGroup{\Sp}_{2n} \to \algebraicGroup{\GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\Sp}_{2n}$  &$\algebraicGroup{\SO}_{2n+1}\to \algebraicGroup{\GL}_{2n+1}$\tabularnewline \hline		 
		$\algebraicGroup{\SO}^{+}_{2n}$ &$\algebraicGroup{\SO}^{+}_{2n}\to \algebraicGroup{\GL}_{2n}$\tabularnewline \hline
		$\algebraicGroup{\SO}^{-}_{2n}$ &$\algebraicGroup{\SO}^{-}_{2n}\to \algebraicGroup{\GL}_{2n}$\tabularnewline \hline
		$\algebraicGroup{\GSp}_{2n}$  &$\algebraicGroup{GSpin}_{2n+1}\to \algebraicGroup{\GL}_{2n+1}$ \tabularnewline \hline
		$\algebraicGroup{\GSO}^{+}_{2n}$ & $\algebraicGroup{GSpin}_{2n}^{+}\to\algebraicGroup{\GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\GSO}^{-}_{2n}$& $\algebraicGroup{GSpin}_{2n}^{-}\to\algebraicGroup{\GL}_{2n}$\tabularnewline \hline
	\end{tabular}
\end{center}
In each of the cases in the table, the map on cocharacters $\rho_{\ast}:\CocharacterLattice{\algebraicGroup{T}^{\ast}} \cong (\mathbb{Q}/\mathbb{Z})_{p'}^n \to \CocharacterLattice{\tilde{\algebraicGroup{T}}^{\ast}} \cong (\mathbb{Q}/\mathbb{Z})_{p'}^{2n}$ sends $x$ to $(x,-w_n x)$. \calvin{in the GSpin cases might scale up by 2}
%Note that in the $SO$ and $GSO$ case, the map $\rho$ sends the split torus of $G^*$ into the split torus of $\GL(2n)$ and the induced map of cocharacters sends $x \mapsto (x,-x)$. In the $Sp,GSp$ cases the map is similar: the induced map of cocharacters sends $x \mapsto (x,1,-x)$. In the unitary case, the torus whose $\finiteField$ points are $(\NormOneGroup{2})^n \subset G^*(\finiteField)$ includes into the torus in $\GL(2n)$ whose $\finiteField$ points are $(\finiteFieldExtension{2}^\times)^n$ and the induced map of cocharacters is $(x_1,x_2, \hdots, x_n) \mapsto (x_1,-x_1,x_2,-x_2, \hdots ,x_n,-x_n)$.

>>>>

The main result of this paper is 
\begin{theorem}
\label{thm:Phi_stable}
	The assignments $H \to \cComplex$ given by $$h \mapsto \genHermitianJacobiKernel{\hermitianSpace}{\chi}\left(h\right)$$ are stable functions. Equivalently, the assignments $$\pi \mapsto \dblJacobiSumScalar{\pi}{\chi}$$ are constant on geometric Lusztig series.
\end{theorem}
This result will be proved in \Cref{subsec:stability}. Our next goal is to prove the following intermediate result.
\begin{lemma}
\label{lem:indep_geo_conj}
    If $(\algebraicGroup{T}_H,\theta_H)$ and $(\algebraicGroup{T}'_H, \theta_H)$ are geometrically conjugate, then $$\frac{\dblVirtualJacobiSumScalar{\RTTheta{T_H}{\theta_H}}{\chi}}{\RTTheta{T_H}{\theta}(1)} = \frac{\dblVirtualJacobiSumScalar{\RTTheta{T'_H}{\theta'}}{\chi}}{\RTTheta{T'_H}{\theta'_H}\left(1\right)}.$$
\end{lemma}

We recall that \elad{Maybe we can help the reader and tell them that $m=1,2$} $$
	\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N} = \prod_{a \in \mathbb{Z}/m\mathbb{Z}} \left(\algebraicGroup{\GL}_N\right)_a,
$$
with Frobenius root action $$\Frobenius(g_0,g_1, \hdots, g_{m-1}) = (\Frobenius(g_{m-1}), \Frobenius(g_0), \Frobenius(g_1), \hdots, \Frobenius(g_{m-2})).$$ We will freely identify $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}(\finiteField) = \GL_N(\finiteFieldExtension{m})$ via the map $\left(g_0,\dots,g_{m-1}\right) \mapsto g_0$.

\begin{remark}
Torus character pairs for $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}$ are the same as torus character pairs for $\algebraicGroup{\GL}_N$ as a group over $\finiteFieldExtension{m}$.
\end{remark}

@@

Given a torus character pair $(\algebraicGroup{T},\theta)$ with associated $\Frobenius$-twisted conjugacy class $(w) \in W(\algebraicGroup{T}_0)$ along with an embedding $\rho:\algebraicGroup{G}^* \to \restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}$ and a Frobenius stable maximal torus $\algebraicGroup{T}^* \subset \restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}$ such that $\rho(\algebraicGroup{T}_0^*) \subset \algebraicGroup{T}^*$, let $\tilde{\algebraicGroup{T}}$ be the torus of $\algebraicGroup{\GL}_N$ \elad{of $\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}$?} corresponding to the $\Frobenius$-twisted conjugacy class of $\rho(w) \in W(\algebraicGroup{T})$ \elad{of $W\left(\algebraicGroup{T}^{\ast}\right)$? Or maybe $W\left(\restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{A_N}\right)$?} and let $\tilde{\theta} \colon \tilde{\algebraicGroup{T}}^{\Frobenius} \to \multiplicativegroup{\cComplex}$ be a character such that $(\rho(s_{\algebraicGroup{T},\theta})) = (s_{\tilde{\algebraicGroup{T}},\tilde{\theta}})$. We define
\[
    g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi) := \tau_{\Tilde{\algebraicGroup{T}}}(\tilde{\theta} \times \chi \circ \det,\psi) = \epsilon_{\Tilde{\algebraicGroup{T}}} q^{-nm/2}\sum_{t \in \tilde{\algebraicGroup{T}}^{\Frobenius}} \tilde{\theta}^{-1}(t)\chi^{-1}({\det}_{\finiteFieldExtension{m}}t) \psi(\trace_{\finiteFieldExtension{m}}t).
\]
Here $\chi \colon \multiplicativegroup{\finiteFieldExtension{m}} \to \multiplicativegroup{\cComplex}$ and $\fieldCharacter \colon \finiteFieldExtension{m} \to \multiplicativegroup{\cComplex}$ are characters. The following lemma shows that this is well defined.

\begin{lemma}
\label{lemma:gl_invariance}
\begin{enumerate}[(i)]
	\item \label{item:gl-n-hasse-davenport-gauss-sum-implies-well-defined} In the case that $\rho$ is the identity map, $g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi)$ depends only on the semisimple conjugacy class of $(s_{\algebraicGroup{T},\theta})$ and the character $\chi$.
	\item The value $g_{\algebraicGroup{T},\rho}(\chi,\theta,\psi)$ is well defined, i.e., it is independent of the choices $\algebraicGroup{T}_0$, $\algebraicGroup{T}^*$ and of the conjugacy class representative $w$.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}[(i)]
	\item In this case when $T \cong \multiplicativegroup{\finiteFieldExtension{\lambda}}$, we recover the Gauss sum
	\[
	g_{T,\rho}(\chi,\theta,\psi) = \GaussSumTorusCharacter{T}{\theta}{\chi}{\fieldCharacter} = \left(-1\right)^n \GaussSumScalar{\tau \times \chi}{\fieldCharacter},
	\]
	where $\tau$ is any irreducible subrepresentation of $\RTThetaVirtual{T}{\GL_N\left(\finiteField\right)}{\theta}$. By \Cref{thm:KondoClassical}, this only depends on the geometric conjugacy class of $(T,\theta)$ and hence only on the semisimple conjugacy class of $(s_{T,\theta})$.
	\item For a Frobenius stable semisimple conjugacy class $(s)$ in $\GL_N$, we define
	$$
	g_{(s)}(\chi,\fieldCharacter) := g_{T,\rho}(\chi,\theta,\fieldCharacter)$$
	where $(T,\theta)$ is a torus character pair such that $\left(s_{T,\theta}\right) = (s)$.
	By (\ref{item:gl-n-hasse-davenport-gauss-sum-implies-well-defined}) this is well defined. Then
	$$
	g_{T,\rho}(\chi,\theta,\fieldCharacter)  = g_{(\rho(s_{T,\theta}))}(\chi,\fieldCharacter),
	$$
	which is independent of the choices $\algebraicGroup{T}^* \subset \algebraicGroup{G}^*$ and $\algebraicGroup{T} \subset \algebraicGroup{\GL}_N$.
\end{enumerate}
\end{proof}

@@

The idea behind the proof of \Cref{lem:indep_geo_conj} is that the formula from \Cref{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} relates $\dblVirtualJacobiSumScalar{\RTGTheta{T_H}{H}{\theta_H}}{\chi}/\RTGTheta{T_H}{H}{\theta_H}\left(1\right)$ to $g_{T,\rho}(\chi,\theta,\psi)$ for the appropriate \calvin{doubling representation} $\rho$ in the table below.

\begin{center}
	\begin{tabular}{|c|c|} \hline
		$\algebraicGroup{G}$ & $\rho \colon \algebraicGroup{G}^{\ast} \to \restrictionOfScalars{\finiteFieldExtension{m}}{\finiteField}{\algebraicGroup{\GL}_N}$  \tabularnewline \hline \hline
		$\algebraicGroup{GL}_n$ &  $\algebraicGroup{\GL}_n \to \algebraicGroup{GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\UnitaryGroup}_{n}$ &  $\algebraicGroup{\UnitaryGroup}_{n} \to \restrictionOfScalars{\quadraticExtension}{\finiteField}{\algebraicGroup{GL}_n}$\tabularnewline \hline		 				 
		$\algebraicGroup{\SO}_{2n+1}$ &$\algebraicGroup{\Sp}_{2n} \to \algebraicGroup{\GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\Sp}_{2n}$  &$\algebraicGroup{\SO}_{2n+1}\to \algebraicGroup{\GL}_{2n+1}$\tabularnewline \hline		 
		$\algebraicGroup{\SO}^{+}_{2n}$ &$\algebraicGroup{\SO}^{+}_{2n}\to \algebraicGroup{\GL}_{2n}$\tabularnewline \hline
		$\algebraicGroup{\SO}^{-}_{2n}$ &$\algebraicGroup{\SO}^{-}_{2n}\to \algebraicGroup{\GL}_{2n}$\tabularnewline \hline
		$\algebraicGroup{\GSp}_{2n}$  &$\algebraicGroup{GSpin}_{2n+1}\to \algebraicGroup{\GL}_{2n+1}$ \tabularnewline \hline
		$\algebraicGroup{\GSO}^{+}_{2n}$ & $\algebraicGroup{GSpin}_{2n}^{+}\to\algebraicGroup{\GL}_{2n}$ \tabularnewline \hline
		$\algebraicGroup{\GSO}^{-}_{2n}$& $\algebraicGroup{GSpin}_{2n}^{-}\to\algebraicGroup{\GL}_{2n}$\tabularnewline \hline
	\end{tabular}
\end{center}
In each of the cases in the table, $m = 1$, except for the unitary group case where $m=2$. In each case, the map on cocharacters $\rho_{\ast}:\CocharacterLattice{\algebraicGroup{T}^{\ast}} \cong (\mathbb{Q}/\mathbb{Z})_{p'}^n \to \CocharacterLattice{\tilde{\algebraicGroup{T}}^{\ast}} \cong (\mathbb{Q}/\mathbb{Z})_{p'}^{2n}$ sends $x$ to $(x,-w_n x)$. \calvin{in the GSpin cases might scale up by 2}
%Note that in the $SO$ and $GSO$ case, the map $\rho$ sends the split torus of $G^*$ into the split torus of $\GL(2n)$ and the induced map of cocharacters sends $x \mapsto (x,-x)$. In the $Sp,GSp$ cases the map is similar: the induced map of cocharacters sends $x \mapsto (x,1,-x)$. In the unitary case, the torus whose $\finiteField$ points are $(\NormOneGroup{2})^n \subset G^*(\finiteField)$ includes into the torus in $\GL(2n)$ whose $\finiteField$ points are $(\finiteFieldExtension{2}^\times)^n$ and the induced map of cocharacters is $(x_1,x_2, \hdots, x_n) \mapsto (x_1,-x_1,x_2,-x_2, \hdots ,x_n,-x_n)$.
